Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions

### 2025-08-28 paper.tex
Making edits

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning Network Growth Mechanisms from Degree Distributions in Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, discrete extremes, power law, preferential
attachment},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Learning Network Growth Mechanisms from Degree Distributions in
Preferential Attachment Networks]{Learning Network Growth Mechanisms
from Degree Distributions in Preferential Attachment Networks}

@@

Networks have become very powerful tools for representing and analysing
complex systems, with uses in a large array of fields, from using
stochastic block models to detect communities online \citep{Latouche11},
to using Exponential Random Graph Models (ERGMs) to analyse the global
trade network \citep{Setayesh22}, and even using mechanistic models to
investigate the patterns in neural systems \citep{Betzel17}.

Along with the recent rise of interest in networks, has been a debate
amongst those that study them regarding the claims that most real
networks are scale-free. This claim is most often equivalent to claiming
that the degree distributions of real networks follow a power law, that
is the fraction of nodes with degree \(k\) is proportional to
\(k^{-\alpha}\), and therefore has a regularly varying tail with extreme
value index \(1/\alpha\). \citet{Broido_2019} provide empirical
evidence, using almost one thousand networks, that these scale-free
networks do not make up a large majority of real networks. They compare
the fits of a power law model against that of several non-scale-free
models only finding strong evidence for scale-freeness in four percent
and weak evidence in fifty-two percent of the networks. On the other
side of this debate is \citet{Voitalov_2019} who disagrees and claims
that these networks are not nearly as rare and only appear to so as a
result of an unrealistic expectation of a power law without deviations
or noise. Evidence of these deviations from a power law is shown by
\citet{Lee24} who demonstrate that a lot of networks are partially
scale-free, in that the body of the degree distribution is modelled well
by a power law while the tail is lighter than what is implied by the
body (although still regularly varying).

Most of the study into the degrees of real networks, including those
mentioned here, has been purely descriptive in the sense that no
information about the networks growth is revealed (aside from relating a
power law to preferential attachment).

The study of network growth began to gain popularity when
\citet{Barabasi99} showed that a network that has grown in such a way
that nodes gain edges at a rate proportional to their degree results in
a network that is scale-free. This is why the presence of real
scale-free networks is such a hot topic, since if it is shown that they
are often scale-free then a model such as preferential attachment gains
a lot of justification for being used to model network growth. The
preferential attachment model they introduced uses a preference
function, \(b(k)=k+1\) where \(k\) is a node in-degree, that is fairly
simple and provided the foundations for many modifications and
extensions to the model. \citet{krapivsky01} considered a more general
preference function of the form \(b(k) = (k+1)^\alpha\) and showed that
the degree distribution only has regular variation in the tail when
\(\alpha=1\), no regular variation when \(0<\alpha<1\) (and therefore
not power law), and when \(\alpha>1\) a finite number of nodes end up
with all edges after a certain point resulting in a degenerate degree
distribution. \citet{wang2022random} returns to a linear preference
function of the form \(b(k) = k+\varepsilon\) but adds the possibility
for reciprocal edges to be sent, this results in joint in-degree
out-degree distribution that is multivariate regular varying and has the
property of hidden regular variation. \citet{rudas07} follows in the
footsteps of \citet{krapivsky01} and considers a preferential attachment
tree with a preference function \(b(\cdot)\) and uses theory from
continuous time branching processes to derive a limiting degree
distribution in terms of the preference function \(b(\cdot)\). Research
in this area tends to only focus on the theoretical asymptotic results
of network growth models with little analysis of real networks. This
paper aims to address this gap, asking if a network is assumed to come
from a preferential attachment model can we use the degree distribution
alone to directly infer the model parameters?

It is important to think about how we intend to consider the tail of the
degree distributions, because if not done correctly we may end up
essentially discounting the effects of the largest degrees (often the
most influential nodes) deviating from a power law. For this reason we
will use methods from discrete extreme value theory, as they best
capture the nature of the data involved, more specifically we will use
results from \citet{shimura12} to study how the tail of the degree
distribution is affected by the preference function since we wish to
align what we do with the recent findings that the a lot of real degree
distributions are regularly varying.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the preferential attachment model alongside the
theoretical results for the survival of the limiting degrees with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). It is shown that a regularly varying degree distribution
can only be the result of an asymptotically linear preference function.
This section also introduces a preference function that can guarantee
regular variation in the degrees while remaining flexible via a set of
parameters that can provide behaviour similar to that in
\citet{krapivsky01} up until a threshold. Section~\ref{sec-model}
utilises the preference function proposed at the end of
Section~\ref{sec-tail}, and illustrates how the extreme value index
(EVI) of the degree distribution varies with the model parameters.
Section~\ref{sec-sim} consists of a simulation study where networks are
simulated from the proposed model for various parameter combinations,
demonstrating that the parameters can be recovered from fitting the
model to only the degree distribution. Section~\ref{sec-real} fits the
model to some real data and provides posterior estimates for the
preference function. Section~\ref{sec-conc} concludes the article.

@@

Starting at time \(t=0\) with an initial network of \(m\) vertices that
each have no edges, at times \(t=1,2,\ldots\) a new vertex is added to
the network bringing with it \(m\) directed edges (with the new vertex
as the source); the target for each of these edges are selected from the
vertices already in the network with weights proportional to some
preference function \(b(\cdot)\) of their degree, where
\(b: N \mapsto \mathbb R^+\setminus\{0\}\) is such that:

@@

Special cases of this model include the Barabási-Albert (BA) model when
\(b(k) = k+\varepsilon\), which leads to a power-law degree distribution
with EVI \(\xi=1/2\), and the Uniform Attachment (UA) model where
\(b(k)=c\) leading to a degree distribution that is not regularly
varying.

The survival function of the limiting degree distribution, called the
limiting survival hereafter, under condition \ref{eq-condb2} can be
analytically derived in the case where \(m=1\).

Consider a continuous time branching process \(\zeta(t)\) driven by a
Markovian pure birth process, with \(\zeta(0)=0\) abd birth rates
depending on a non-negative function \(b(\cdot)\):

@@

Theorem 1 from \citet{rudas07} states that for the tree \(\Upsilon(t)\)
at time \(t\):

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]dt
}\end{equation} where \(\lambda^*\) satisfies \(\hat\rho(\lambda^*)=1\)
and \(\hat\rho\) is the Laplace transform of the density of the point
process associated with the pure birth process that corresponds to a
nodes individual growth, that is:

\[
\hat\rho(\lambda) \coloneq \int_0^\infty e^{-\lambda t}\rho(t)\mathrm{d}t.
\]

@@

\[
\bar F(k) = \frac{\int_0^\infty e^{-\lambda^* t}\mathbb E\left[\mathbb I\left\{\text{deg}(x,\Upsilon(t))>k\right\}\right]dt}{\int_0^\infty e^{-\lambda^* t}dt} = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}.
\] Therefore, the corresponding probability mass function of the degree
distribution \(f(k) = \bar F(k-1) - \bar F(k)\) is

\[
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)}.
\]

@@

For a distribution \(F\) with survival function \(\bar F\) and some
\(k\in\mathbb Z^+\) let

@@

\citet{shimura12} states that if
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is regularly varying with \(\bar F(k) \sim k^{-\alpha}\). On
the other hand, if \(\lim_{k\rightarrow\infty} \Omega(F,k) = 0\) then we
will refer to the distribution as light-tailed. This allows us to show
the following:

@@

Proposition~\ref{prp-omega} aligns with the result from
\citet{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light-tailed distribution, as
\(\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0\) if \(b(k)=k^\alpha\) where
\(\alpha < 1\). Proposition 2.1 also aligns with the fact that BA model
produces a regularly varying degree distribution with EVI \(\xi=0.5\) by
considering the preference function \(b(k) = k + \varepsilon\), as
\(\lim_{k\rightarrow\infty}b(k+1)-b(k)=1\) leaving the tail index to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\). So,
in order for the degree distribution to be regularly varying we need
that the limit \(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is
positive. To determine the class of functions that will result in
regularly varying limiting degree distributions, we use the following
result:

@@

Consider a GPA model with preference function \(b(\cdot)\) satisfying
Proposition~\ref{prp-omega}. Then the limiting degree distribution is
regularly varying with EVI \(c/\lambda^{*}\) if and only if
\(\lim_{k\rightarrow\infty}b(k)/k = c > 0\).

@@

\[
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
\] meaning by Theorem~\ref{thm-stolz} in Appendix \ref{sec-sup}:

@@

Using this result we can understand how the preference function is
directly connected with the EVI and regular variation of the degree
distribution. We use this result to create a preference function that
guarantees regular variation in the tail of the degree distribution,
aligning with analysis of real networks, whilst allowing for the tail to
deviate from the shape implied by the body. This gives the model the
capability to produce realistic behaviour in the degrees like what was
in \citet{Lee24} by using a piecewise function inspired by the observed
deviation from the power law after a certain threshold.

@@

This preference function, as per Proposition~\ref{prp-omega}, will
produce a degree distribution with EVI \(\xi=\beta/\lambda^*\)
guaranteeing regular variation. We study this preference function
further in the next section.

@@

In the previous section, we found that using a preference function with
linearity in the limit allows for the inclusion of sub/super-linear
behaviour below the threshold, while simultaneously guaranteeing regular
varation of the degrees. In this section, we study how the limiting
degree distribution varies with the preference function according to
Equation~\ref{eq-pref}. Its survival is

@@

which must be solved numerically for most parameter choices. Also, note
that \(\lambda>\beta\).

Some examples of the limiting degree distribution for various parameter
combinations are shown below on log-log scale in
Figure~\ref{fig-polylinsurv}:

@@

\caption{\label{fig-polylinsurv}Theoretical survival distributions of
the limiting degree distributions, according to various combinations of
\((\alpha, \beta, \varepsilon)\) and \(k_0=20\) of the proposed
preferential attachment model.}

@@

Figure~\ref{fig-polylinsurv} demonstrates that this model can capture a
wide variety of shapes for the survival function, including a large
range of possible tail indices ranging from 0.035
(\(\alpha=1.5, \beta=0.1, \varepsilon=1\)) to 0.999
(\(\alpha=0.5, \beta=1.5, \varepsilon=0.01\)).

The survival function (\ref{eq-polysurv}) can be connected to the
discrete version of the generalised Pareto distribution (GP), called the
Integer GP (IGP) \citep{Rohrbeck_2018} with conditional survival: \[
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
\] for \(v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R\), denoted as
\(X|X>u \sim  \mathrm {IGP}(\xi, \sigma, u)\) where \(\xi\) is the shape
parameter controlling the tail index

@@

\begin{align*}
\bar F(k|k\ge k_0) &= \frac{\Gamma\left(\frac{\lambda^* + k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(\frac{k_0^\alpha + \varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0  +1 + \frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0  +1 + \frac{\lambda^*+ k_0^\alpha + \varepsilon}{\beta}\right)}\\
&\approx\left(\frac{k_0^\alpha+\varepsilon}{\beta}\right)^{\lambda^*/\beta}\left(k-k_0+1+\frac{k_0^\alpha + \varepsilon}{\beta}\right)^{-\lambda^*/\beta}\\
&=\left(\frac{k_0^\alpha+\varepsilon}{k_0^\alpha+\varepsilon + \beta}\right)^{\lambda^*/\beta}\left(\frac{\beta(k-k_0)}{\beta + k_0^\alpha+\varepsilon} + 1\right)^{-\lambda^*/\beta}\\
&=\left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^{*}/\beta}
\end{align*}

Therefore, \[
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
\] meaning that for \(k\ge k_0\) the limiting degree distribution (for
large \(k_0^\alpha\)) is approximated by
\(\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)\).

To assess how close of an approximation this is, the theoretical
conditional survivals are shown in Figure~\ref{fig-approx_surv} in
colour and their IGP approximations are shown in grey. The approximation
seems to hold up fairly well even for large degrees and more so when
\(\alpha\) is larger.

@@

In agreement with Proposition~\ref{prp-omega2}, \(\beta>0\) ensures that
the shape parameter of the IGP is positive and thus the distribution is
regularly varying. Additionally the value of the shape parameter \(\xi\)
is shown in Figure~\ref{fig-polyheat} for various parameter choices. The
darker and lighter regions on the heat maps correspond to a heavier and
a lighter tail, respectively, the red dashed line shows combinations of
\(\alpha\) and \(\beta\) that produce a limiting degree distribution
with the same tail index as the Barabási-Albert model, that is,
\(\xi=0.5\).

Through the connection to the IGP, fitting this model is almost
equivalent to fitting the IGP to the degree and estimating the
parameters but instead of only describing the shape of the degree
distribution we would also gain estimates for the shape of the
preference function while can help understand the mechanisms underlying
the growth of the network.

@@

\caption{\label{fig-rec1}Posterior estimates of survival function for
data simulated from the proposed model with various combinations of
(\(\alpha\),\(\beta\),\(\varepsilon\)) and \(k_0=20\).}

>>>>

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, discrete extremes, power law, preferential
attachment},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Learning growth mechanisms of tail realistic preferential
attachment models from network degree distributions]{Learning growth
mechanisms of tail realistic preferential attachment models from network
degree distributions}

@@

Networks have become very powerful tools for representing and analysing
complex systems, with uses in a large array of fields. In network
science and statistics, they have been studies by various families of
models, from stochastic block models for detecting communities online
\citep{Latouche11}, to exponential random graph models (ERGMs) for
analysing the global trade network \citep{Setayesh22}, and mechanistic
models for investigating patterns in neural systems \citep{Betzel17}.

Amid the recent rise of interest in networks, there has been a debate on
whether most real networks are scale-free. Claiming a real network is
scale-free is equivalent to saying that its degree distribution follows
a power law, that is the fraction of nodes with degree \(k\) is
proportional to \(k^{-\alpha}\), and therefore has a regularly varying
tail with tail index \(\alpha\). On the side against the claim is
\citet{Broido_2019who} compared the fits of a power law model against
that of several non-scale-free models to nearly a thousand networks,
only to find strong evidence for scale-freeness in four percent and weak
evidence in over half of the networks, thus claiming that scale-free
networks do not make up a majority in real networks. They compare the
fits of a power law model against that of several non-scale-free models
only finding strong evidence for scale-freeness in four percent and weak
evidence in fifty-two percent of the networks. On the other side of this
debate is \citet{Voitalov_2019} who disagrees and claims that these
networks are not nearly as rare and only appear to so as a result of an
unrealistic expectation of a power law without deviations or noise.
Additional evidence of these deviations from a power law is shown by
\citet{Lee24} who demonstrate that a lot of networks are partially
scale-free, in that the body of the degree distribution is often
modelled well by a power law, while the tail is lighter than what is
implied by the body, albeit still regularly varying..

Nevertheless, most studies into the appropriateness of a power law for
the degrees of real networks, the aforementioned references included,
have been largely descriptive in the sense that no information about the
growth of the networks is revealed.

The popularity of using the power law for network degrees can be traced
back to the preferential attachment (PA) model popularised by
\citet{Barabasi99}. In the general model, as new nodes join the network,
an existing node with degree k gains edges at a rate proportional to
\(b(k)\), where b is a non-negative preference function.
\citet{Barabasi99} showed that, when \(b(k) = k + 1\), in the limit the
resulting degree distribution is regularly varying with index 3.
Subsequently, if a real network is shown to be scale-free, one can
loosely justify preferential attachment as the underlying mechanism of
its growth.

The model from \citet{Barabasi99} provided the foundations for various
generalisations.

\citet{krapivsky01} considered \(b(k) = (k+1)^\alpha\), and showed that
the degree distribution is not regularly varying (and therefore not
following the power law) when \(0<\alpha<1\), and when \(\alpha>1\) a
finite number of nodes end up with all edges after a certain point
resulting in a degenerate degree distribution. \citet{wang2022random}
returns to a linear preference function of the form
\(b(k) = k+\varepsilon\) but adds the possibility for reciprocal edges
to be sent, resulting in the joint distribution of in-degree and
out-degree being multivariate regularly varying and having the property
of hidden regular variation. \citet{rudas07} follows in the footsteps of
\citet{krapivsky01} , by considering a preferential attachment tree and
using theory from continuous branching processes, derives a limiting
degree distribution in terms of the preference function \(b(\cdot)\).
Nevertheless, research in this area tends to only focus on the
theoretical asymptotic results of network growth models with little
analysis of real networks.

This paper aims to address the gap between the applied and theoretical
works, by asking if a network is assumed to come from a preferential
attachment model, can we use the degree distribution alone to directly
infer the parameters of the the preference function and learn about the
growth mechanisms? Moreover, proper consideration is given to the tail
of the degree distribution, because otherwise the effects of the largest
degrees, which correspond to the most influential nodes, deviating from
the power law will be discounted.

It is important to think about how we intend to consider the tail of the
degree distributions, because if not done correctly we may end up
essentially discounting the effects of the largest degrees (often the
most influential nodes) deviating from a power law.

For the above reason, we will use methods from discrete extreme value
theory, in particular those by \citet{shimura12}, who provided
theoretical guarantees for a discrete distribution to be regularly
varying or not. Using these results, we demonstrate how the tail of the
degree distribution is affected by \(b(\cdot)\), and subsequently
propose a class of preference functions that is tail realistic for real
networks. These analytical results enables the likelihood of the degree
distribution to be expressed in terms of the parameters of \(b(\cdot)\),
which in turn allows the underlying mechanism of the network, assumed to
grow according to preferential attachment, to be inferred directly.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the preferential attachment model alongside the
theoretical results for the survival of the limiting degrees, with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). A class of asymptotically linear preference functions will
be introduced and shown to guarantee regular variation in the degrees
while remaining flexible up until a threshold. Section~\ref{sec-model}
utilises the proposed preference function and illustrates numerically
how the tail index of the degree distribution varies with the model
parameters. Section~\ref{sec-sim} consists of a simulation study where
networks are simulated from the proposed model for various parameter
combinations, demonstrating that the parameters can be recovered from
fitting the model to only the degree distribution.
Section~\ref{sec-real} fits the model to some real data and provides
posterior estimates for the preference function. Section~\ref{sec-conc}
concludes the article.

@@

Starting at time \(t=0\) with an initial network of \(m\) vertices that
each have no edges, at times \(t=1,2,\ldots\) a new vertex is added to
the network bringing with it \(m\) directed edges from the new vertex;
the target for each of these edges are selected from the vertices
already in the network with weights proportional to some non-decreasing
preference function \(b(\cdot)\) of their degree, where
\(b: \mathbb N \mapsto \mathbb R^+\setminus\{0\}\) is such that:

@@

Special cases of this model include the Barabási-Albert (BA) model when
\(b(k) = k+\varepsilon\), which leads to a power-law degree distribution
with tail index 2, and the Uniform Attachment (UA) model where
\(b(k)=c\) leading to a degree distribution that is not regularly
varying.

The survival function of the limiting degree distribution, called the
limiting survival hereafter, under condition \ref{eq-condb2} can be
analytically derived in the case where \(m=1\), this is presented below.

Consider a continuous time branching process \(\zeta(t)\) driven by a
Markovian pure birth process, with \(\zeta(0)=0\) and birth rates
depending on a non-negative function \(b(\cdot)\):

@@

Theorem 1 from \citet{rudas07} states that for the tree \(\Upsilon(t)\)
at time \(t\) and a characteristic function of the tree
\(\varphi(\cdot)\) :

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]dt
}\end{equation} where \(\lambda^*\) satisfies \(\hat\rho(\lambda^*)=1\)
and \(\hat\rho\) is the Laplace transform of the density of the point
process associated with the pure birth process that corresponds to the
growth of an individual node, that is
\(\hat\rho(\lambda) \coloneq \int_0^\infty e^{-\lambda t}\rho(t)\mathrm{d}t\)

@@

\begin{equation}\phantomsection\label{eq-surv-origin}{
\bar F(k) = \frac{\int_0^\infty e^{-\lambda^* t}\mathbb E\left[\mathbb I\left\{\text{deg}(x,\Upsilon(t))>k\right\}\right]dt}{\int_0^\infty e^{-\lambda^* t}dt} = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}.
}\end{equation}

Therefore, the corresponding probability mass function of the degree
distribution \(f(k) = \bar F(k-1) - \bar F(k)\) is

\begin{equation}\phantomsection\label{eq-pmf-origin}{
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)}.
}\end{equation}

@@

For a distribution \(F\) with survival function \(\bar F\) and
\(k\in\mathbb Z^+\) let

@@

This allows us to show the following:

@@

\citet{shimura12} states that if
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is regularly varying with \(\bar F(k) \sim k^{-\alpha}\). On
the other hand, if \(\lim_{k\rightarrow\infty} \Omega(F,k) = 0\) then we
will refer to the distribution as light-tailed.

Proposition~\ref{prp-omega} aligns with the result from
\citet{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light-tailed distribution, as
\(\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0\) if \(b(k)=k^\alpha\) where
\(\alpha < 1\). Proposition 2.1 also aligns with the fact that BA model
produces a regularly varying degree distribution, with tail index 2, by
considering the preference function \(b(k) = k + \varepsilon\), as
\(\lim_{k\rightarrow\infty}b(k+1)-b(k)=1\) leaving the tail index to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\). So,
in order for the degree distribution to be regularly varying we need
that the limit \(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is
positive. The following proposition determines the class of functions
that will result in regular varying degree distributions.

@@

Consider a GPA model with preference function \(b(\cdot)\) satisfying
Proposition~\ref{prp-omega}. Then the limiting degree distribution is
regularly varying with tail index \(\lambda^{*}/c\) if and only if
\(\lim_{k\rightarrow\infty}b(k)/k = c > 0\).

@@

\[
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
\] Using the Stolz-Cesaro Theorem
\href{Appendix\%20-@sec-sup}{@thm-stolz},

@@

Using this result we can understand how the preference function is
directly connected with the tail behaviour of the degree distribution.
Specifically, regular variation is achieved if and only if \(b(k)\) is
asymptotically linear with \(k\). We use this result to create a
preference function that guarantees regular variation in the tail of the
degree distribution, aligning with analysis of real networks, whilst
allowing for the tail to deviate from the shape implied by the body.
This gives the model the capability to produce realistic behaviour in
the degrees like what was in \citet{Lee24} by using a piecewise function
inspired by the observed deviation from the power law after a certain
threshold:

@@

Using Proposition~\ref{prp-omega}, we can show that
\(\lim_{n\rightarrow\infty}\Omega(F,k)=\lambda^{*}/\beta\), meaning the
degree distribution obtained using this preference funnction is
regularly varying with tail index \(\lambda^*/\beta\).

@@

In the previous section, we found that using an asymptotically linear
preference function allows for the inclusion of sub/super-linear
behaviour below the threshold, while simultaneously guaranteeing regular
varation of the degrees. In this section, we demonstrate the flexibility
of the preference function in Equation~\ref{eq-pref}, with regard to the
tail behaviour of the limiting degree distribution. Using
Equation~\ref{eq-surv-origin}, the limiting survival is

@@

which has to be solved numerically for most parameter choices. Also,
note that \(\lambda>\beta\).

For some parameter combinations, the limiting survival \(\bar F(k)\) is
shown on log-log scale in Figure~\ref{fig-polylinsurv}:

@@

\caption{\label{fig-polylinsurv}The limiting survival, according to
various combinations of \((\alpha, \beta, \varepsilon)\) and \(k_0=20\)
of the proposed preferential attachment model.}

@@

Figure~\ref{fig-polylinsurv} demonstrates that this model can capture a
range of tail behaviour, including a large range of possible tail
indices ranging from 0.035 (\(\alpha=1.5, \beta=0.1, \varepsilon=1\)) to
0.999 (\(\alpha=0.5, \beta=1.5, \varepsilon=0.01\)).

The analytic form of the survival function in (\ref{eq-polysurv}),
offers a natural connection to the discrete version of the generalised
Pareto distribution (GP), providing a link to a well established
component of discrete extremes in the literature. Specifically,
connecting to the Integer GP (IGP) seen in \citet{Rohrbeck_2018} with
conditional survival: \[
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
\] for \(v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R\), denoted as
\(X|X>u \sim  \mathrm {IGP}(\xi, \sigma, u)\) where \(\xi\) is the shape
parameter controlling the tail index.

@@

\begin{align*}
\bar F(k|k\ge k_0) &= \frac{\Gamma\left(\frac{\lambda^* + k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(\frac{k_0^\alpha + \varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0  +1 + \frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0  +1 + \frac{\lambda^*+ k_0^\alpha + \varepsilon}{\beta}\right)}\\
&\approx\left(\frac{k_0^\alpha+\varepsilon}{\beta}\right)^{\lambda^*/\beta}\left(k-k_0+1+\frac{k_0^\alpha + \varepsilon}{\beta}\right)^{-\lambda^*/\beta}\\
&=\left(\frac{k_0^\alpha+\varepsilon}{k_0^\alpha+\varepsilon + \beta}\right)^{\lambda^*/\beta}\left(\frac{\beta(k-k_0)}{\beta + k_0^\alpha+\varepsilon} + 1\right)^{-\lambda^*/\beta}\\
&=\left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^{*}/\beta}.
\end{align*}

Therefore, \[{#eq-igp-est}
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
\] meaning that for \(k\ge k_0\) the limiting degree distribution (for
large \(k_0^\alpha\)) is approximated by the
\(\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)\).

To assess how close of an approximation this is, the theoretical
conditional survival Equation~\ref{eq-polysurv} are shown in
Figure~\ref{fig-approx_surv} in colour and their IGP approximations
\textbf{?@eq-igp-est} are shown in grey. The approximation seems to hold
up fairly well even for large degrees and more so when \(\alpha\) is
larger.

@@

In agreement with Proposition~\ref{prp-omega2}, \(\beta>0\) ensures that
the shape parameter of the IGP is positive and thus the distribution is
regularly varying. Additionally the value of the shape parameter \(\xi\)
is shown in Figure~\ref{fig-polyheat} for various parameter choices. The
darker and lighter regions on the heat maps correspond to a heavier and
a lighter tail, respectively, the red dashed line shows combinations of
\(\alpha\) and \(\beta\) that produce a limiting degree distribution
with the same tail index as the Barabási-Albert model.

Through the connection to the IGP, fitting this model is almost
equivalent to fitting the IGP to the degrees and estimating its
parameters. However, instead of only describing the shape of the degree
distribution, we would also gain estimates for the shape of the
preference function, thus gaining a direct understanding the mechanisms
underlying the growth of the network.

@@

\caption{\label{fig-rec1}Posterior estimates of survival function for
data simulated from the proposed model with various combinations of
\((\alpha,\beta,\varepsilon)\) and \(k_0=20\).}
